---
title: "testing - Eva"
output: html_document
date: "2023-06-09"
---

# packages & prep 
```{r}
# remove.packages(c("mstate","relsurv"))
# 
# install.packages('relsurv_2.3-1.tar.gz', repos=NULL, type='source')
# install.packages('mstate_0.3.2.tar.gz', repos=NULL, type='source')

library(mstate)
library(relsurv)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(colorspace)
state_cols<-heat_hcl(6, c=c(80,30), l=c(30,90), power=c(1.5,2))[c(5,4,3,2,1,6)]
theme_set(theme_bw())
theme_update(panel.background=element_rect(color="black"))


load("joinpoptab.RData")
dim(joinpoptab)
dimnames(joinpoptab)

# load EBMT dataset
data("ebmt4")
ebmt<-ebmt4
head(ebmt)
dim(ebmt)

names(ebmt)

# add sex and country

set.seed(1234567) 
ebmt$sex=sample(rep(c("male","female"),1140)[1:2279])
ebmt$country=sample(rep(c("Netherlands","Germany","Slovenia"),760)[1:2279])

table(ebmt$agecl) # 551 <=20, 1213 20-40, 515 >40

ebmt$age2<-NA
ebmt$age2[ebmt$agecl=="<=20"]<-sample(rep(c(18,19,20),184)[1:551])
ebmt$age2[ebmt$agecl=="20-40"]<-sample(rep(c(21:40),61)[1:1213])
ebmt$age2[ebmt$agecl==">40"]<-sample(rep(c(41:65),21)[1:515])
ebmt%>%group_by(agecl)%>%summarise(n=n(),min(age2),max(age2))

table(ebmt$year) # 634 1985-1989, 896 1990-1994, 749 1995-1998
dimnames(joinpoptab)$year # 1990-2011
# no data available before 1990 -> set years <1990 at 1990...
ebmt$year2<-NA
ebmt$year2[ebmt$year=="1985-1989"]<-1990
ebmt$year2[ebmt$year=="1990-1994"]<-sample(rep(c(1990:1994),180)[1:896])
ebmt$year2[ebmt$year=="1995-1998"]<-sample(rep(c(1995:1998),752)[1:749])
ebmt%>%group_by(year)%>%summarise(n=n(),min(year2),max(year2))

# Tstart, Tstop and time are in days
```

# multi-state model without relative survival
```{r}
tmat<-transMat(list(
  c(2,3,5,6),
  c(4,5,6),
  c(4,5,6),
  c(5,6),
  c(),
  c()),
  names=c("Tx","Rec","AE","Rec+AE","Rel","Death"))

tmat

msebmt<-msprep(data=ebmt, trans=tmat, 
               time=c(NA,"rec","ae","recae","rel","srv"),
               status=c(NA,"rec.s","ae.s","recae.s","rel.s","srv.s"),
               keep=c("match","proph","year","agecl",
                      "age2","year2","sex","country"))
head(msebmt,n=20)
events(msebmt)

msebmt<-expand.covs(msebmt,c("match","proph","year","agecl"),longnames = F)
head(msebmt,n=1)

# non-parametric model
c0<-coxph(Surv(Tstart,Tstop,status)~strata(trans), data=msebmt,method="breslow")
msf0<-msfit(c0, vartype = "greenwood", trans=tmat)
pt0<-probtrans(msf0, predt=0, method="greenwood")
plot(pt0,use.ggplot = T,type="separate")+scale_color_manual(values=rep("black",6))+
  theme(legend.position="none")
plot(pt0,use.ggplot = T,type="filled", ord=c(2,4,3,5,6,1),cols=state_cols)

# from different states at time 100:
pt100<-probtrans(msf0,predt=100)

grid.arrange(plot(pt100,from=1,ord=c(2,4,3,5,6,1),use.ggplot = T)+
               scale_fill_manual(values=state_cols,drop=F)+ggtitle("in Tx at 100 days"),
             plot(pt100,from=3,ord=c(2,4,3,5,6,1),use.ggplot = T)+
               scale_fill_manual(values=state_cols,drop=F)+ggtitle("in AE at 100 days"),
             nrow=1)

# semi-parametric model
cfull<-coxph(Surv(Tstart,Tstop,status)~match.1 + match.2 + match.3 + match.4 + match.5 + match.6 + match.7 +
               match.8 + match.9 + match.10 + match.11 + match.12 + proph.1 + proph.2 + proph.3 + proph.4 +
               proph.5 + proph.6 + proph.7 + proph.8 + proph.9 + proph.10 + proph.11 + proph.12 + year1.1 +
               year1.2 + year1.3 + year1.4 + year1.5 + year1.6 + year1.7 + year1.8 + year1.9 + year1.10 +
               year1.11 + year1.12 + year2.1 + year2.2 + year2.3 + year2.4 + year2.5 + year2.6 + year2.7 + 
               year2.8 + year2.9 + year2.10 + year2.11 + year2.12 + agecl1.1 + agecl1.2 + agecl1.3 + 
               agecl1.4 + agecl1.5 + agecl1.6 + agecl1.7 + agecl1.8 + agecl1.9 + agecl1.10 + agecl1.11 +
               agecl1.12 + agecl2.1 + agecl2.2 + agecl2.3 + agecl2.4 + agecl2.5 + agecl2.6 + agecl2.7 +
               agecl2.8 + agecl2.9 + agecl2.10 + agecl2.11 + agecl2.12 + strata(trans),
             data=msebmt, method = "breslow")
summary(cfull)

whA<-which(msebmt$proph=="yes"&msebmt$match=="no gender mismatch"&msebmt$year=="1995-1998"&
             msebmt$agecl=="<=20")
patA<-msebmt[rep(whA[1],12),9:12]
patA$trans<-1:12
attr(patA,"trans")<-tmat
patA<-expand.covs(patA,covs=c("match","proph","year","agecl"),longnames = F)
patA$strata<-patA$trans
msfA<-msfit(cfull,patA,trans=tmat)
ptA<-probtrans(msfA,predt=0)

whB<-which(msebmt$proph=="no"&msebmt$match=="gender mismatch"&msebmt$year=="1985-1989"&
             msebmt$agecl==">40")
patB<-msebmt[rep(whB[1],12),9:12]
patB$trans<-1:12
attr(patB,"trans")<-tmat
patB<-expand.covs(patB,covs=c("match","proph","year","agecl"),longnames = F)
patB$strata<-patB$trans
msfB<-msfit(cfull,patB,trans=tmat)
ptB<-probtrans(msfB,predt=0)

grid.arrange(plot(ptA,from=1,ord=c(2,4,3,5,6,1),use.ggplot = T)+
               scale_fill_manual(values=state_cols,drop=F)+ggtitle("patient A"),
             plot(ptB,from=1,ord=c(2,4,3,5,6,1),use.ggplot = T)+
               scale_fill_manual(values=state_cols,drop=F)+ggtitle("patient B"),
             nrow=1)
```

# non-parametric multistate model with relative survival
```{r}
msf0_relsurv<-msfit.relsurv(msfit.obj = msf0, data=msebmt, 
                            split.transitions = c(4,7,10,12),
                            ratetable = joinpoptab,
                            rmap=list(age=age2*365.241, year=year2),
                            time.format="days",var.pop.haz = "bootstrap",B=100)

pt_relsurv<-probtrans(msf0_relsurv, predt=0, direction="forward", method="greenwood")
summary(msf0_relsurv, times=c(100,c(1,2,5,10)*365.241))
summary(pt_relsurv, times=c(100,c(1,2,5,10)*365.241), conf.type="log")

plot(msf0_relsurv, use.ggplot=T)
ggplot(plot(msf0_relsurv, use.ggplot=T)$data%>%filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,ymin=CI_low,ymax=CI_upp,col=type))+
  geom_line()+
  facet_wrap(~from)
ggplot(plot(msf0_relsurv,use.ggplot=T)$data%>%rbind(plot(msf0,use.ggplot=T)$data)%>%
         filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=case_when(grepl("\\.p",trans_name)~"population hazard",
                               grepl("\\.e",trans_name)~"excess hazard",
                               T~"total hazard")),
       aes(x=time,y=Haz,color=type))+
  geom_line()+
  facet_grid(~from)

plot(pt_relsurv, use.ggplot=T, type="separate")+
  scale_color_manual(values=rep("black",10))+theme(legend.position="none")
plot(pt_relsurv, use.ggplot=T)

ggplot(plot(pt0,use.ggplot=T,type="separate")$data%>%mutate(type="normal")%>%
         rbind(plot(pt_relsurv,use.ggplot=T,type="separate")$data%>%mutate(type="relsurv"))%>%
         mutate(state2=ifelse(grepl("Death",state),"Death",as.character(state)),
                death=case_when(grepl("\\.p",state)~"population mortality",
                                grepl("\\.e",state)~"excess mortality",
                                T~"total")),
       aes(x=time,y=prob,col=type,lty=death))+
  geom_step(alpha=0.75)+
  facet_wrap(~state2)+
  guides(color=guide_legend(rev=T))

# split death into different death states
tmat2<-transMat(list(
  c(2,3,5,6),
  c(4,5,7),
  c(4,5,8),
  c(5,9),
  c(),
  c(),
  c(),
  c(),
  c()),
  names=c("Tx","Rec","AE","Rec+AE","Rel","Death after Tx","Death after Rec", "Death after AE",
          "Death after Rec+AE"))

tmat2

msebmt2<-msprep(data=ebmt, trans=tmat2, 
               time=c(NA,"rec","ae","recae","rel","srv","srv","srv","srv"),
               status=c(NA,"rec.s","ae.s","recae.s","rel.s","srv.s","srv.s","srv.s","srv.s"),
               keep=c("match","proph","year","agecl","age2","year2","sex","country"))

events(msebmt)$Frequencies
events(msebmt2)$Frequencies

c2<-coxph(Surv(Tstart,Tstop,status)~strata(trans),data=msebmt2,method="breslow")
msf2<-msfit(c2, vartype = "greenwood", trans=tmat2)
pt2<-probtrans(msf2, predt=0, method="greenwood")
plot(pt2,use.ggplot = T,type="separate")+scale_color_manual(values=rep("black",9))+
  theme(legend.position="none")
plot(pt2,use.ggplot = T,type="filled")

msf2_relsurv<-msfit.relsurv(msfit.obj = msf2, data=msebmt2, 
                            split.transitions = c(4,7,10,12),
                            ratetable = joinpoptab,
                            rmap=list(age=age2*365.241, year=year2),
                            time.format="days",var.pop.haz = "bootstrap",B=100)

pt2_relsurv<-probtrans(msf2_relsurv, predt=0, direction="forward", method="greenwood")
plot(pt2_relsurv, use.ggplot=T, type="separate")+
  scale_color_manual(values=rep("black",13))+theme(legend.position="none")
plot(pt2_relsurv, use.ggplot=T)

ggplot(plot(pt0,use.ggplot=T,type="separate")$data%>%mutate(type="normal")%>%
         rbind(plot(pt2_relsurv,use.ggplot=T,type="separate")$data%>%mutate(type="relsurv"))%>%
         mutate(state2=ifelse(grepl("Death",state),"Death",as.character(state)),
                death=case_when(grepl("\\.p",state)~"population mortality",
                                grepl("\\.e",state)~"excess mortality",
                                T~"total")),
       aes(x=time,y=prob,col=type,lty=death))+
  geom_step(alpha=0.75)+
  facet_wrap(~state2)+
  guides(color=guide_legend(rev=T))
```

# semi-parametric multistate model with relative survival
```{r}
source("functions Eva.R")
summary(cfull)

# Perform coxph.relsurv:
cox.rs <- coxph.relsurv2(Surv(Tstart, Tstop, status)~match.1 + match.2 + match.3 + match.4 + match.5 + 
                          match.6 + match.7 + match.8 + match.9 + match.10 + match.11 + match.12 + proph.1 +
                          proph.2 + proph.3 + proph.4 + proph.5 + proph.6 + proph.7 + proph.8 + proph.9 +
                          proph.10 + proph.11 + proph.12 + year1.1 + year1.2 + year1.3 + year1.4 + year1.5 +
                          year1.6 + year1.7 + year1.8 + year1.9 + year1.10 + year1.11 + year1.12 + year2.1 +
                          year2.2 + year2.3 + year2.4 + year2.5 + year2.6 + year2.7 + year2.8 + year2.9 +
                          year2.10 + year2.11 + year2.12 + agecl1.1 + agecl1.2 + agecl1.3 + agecl1.4 +
                          agecl1.5 + agecl1.6 + agecl1.7 + agecl1.8 + agecl1.9 + agecl1.10 + agecl1.11 + 
                          agecl1.12 + agecl2.1 + agecl2.2 + agecl2.3 + agecl2.4 + agecl2.5 + agecl2.6 +
                          agecl2.7 + agecl2.8 + agecl2.9 + agecl2.10 + agecl2.11 + agecl2.12+strata(trans), 
             data=msebmt, method="breslow",
             split.transitions = c(4,7,10,12), ratetable = joinpoptab, time.format = 'days',
             rmap=list(age=age2*365.241, year=year2))
cox.rs
summary(cox.rs)
class(cox.rs)
# Eva: compare cox.rs with 'normal' Cox model, focus on the split transitions

#### patient A ####
names(patA)
# add year2, age2, country and sex
patA%>%select(year,agecl)%>%unique()
patA$year2<-1996
patA$age2<-19
patA$country<-"Netherlands"
patA$sex<-"female"

#save.image()

msfA_relsurv<-msfit.coxph.relsurv(cox.rs, newdata = patA, trans = tmat)
length(table(msfA_relsurv$Haz$trans)) # 14
msf0_relsurv$trans # 16! why??
c(1:16)[which(!c(1:16)%in%msfA_relsurv$Haz$trans)] 
# transition 6 (Rec -> Rel) and 11 (Rec+AE -> Rel)
summary(cox.rs) # coefficients for 6 and 11 are present
# in msfit.coxph.relsurv: we seem to lose the transitions in function starting at line 17
# because of intermediate states after absorbing transitions??

# Plot predicted hazards:
msfA_relsurv$Haz<-
  left_join(msfA_relsurv$Haz,data.frame(trans=1:length(to.trans2(msf0_relsurv$trans)$transname),
                                        trans_name=to.trans2(msf0_relsurv$trans)$transname))
# to.trans2(msf0_relsurv$trans)$transname because msfA_relsurv misses trans 11 & 12
table(msfA_relsurv$Haz$trans,msfA_relsurv$Haz$trans_name)

ggplot(msfA_relsurv$Haz%>%filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_wrap(~from)

ggplot(msfA_relsurv$Haz%>%
         mutate(type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_wrap(~trans_name)
# what's happening for Rec -> Rel??

ggplot(msfA_relsurv$Haz%>%
         mutate(type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_wrap(~gsub("\\.e|\\.p","",trans_name))

events(msebmt2%>%filter(match=="no gender mismatch"&proph=="yes"&year=="1995-1998"&
                          agecl=="<=20"))$Frequencies
# only 3 patients enter relapse, 1 from Tx and 2 from Rec
msfA_relsurv$Haz%>%filter(trans_name=="Rec -> Rel")%>%mutate(time_diff=time-lag(time))%>%
  mutate(diff=Haz-lag(Haz))%>%#filter(time_diff<0)%>%
  arrange(time)%>%filter(diff!=0)%>%
  print()%>%filter(diff<0)

# probabilities
ptA_relsurv<-probtrans(msfA_relsurv, predt=0, variance=F)
plot(ptA_relsurv, use.ggplot=T, type="separate")+
  scale_color_manual(values=rep("black",10))+theme(legend.position="none")
plot(ptA_relsurv, use.ggplot=T)

#### patient B ####
names(patB)
# add year2, age2, country and sex
patB%>%select(year,agecl)%>%unique()
patB$year2<-1990
patB$age2<-60
patB$country<-"Germany"
patB$sex<-"male"
msfB_relsurv<-msfit.coxph.relsurv(cox.rs, newdata = patB, trans = tmat)

# Plot predicted hazards:
msfB_relsurv$Haz<-
  left_join(msfB_relsurv$Haz,data.frame(trans=1:length(to.trans2(msfB_relsurv$trans)$transname),
                                        trans_name=to.trans2(msfB_relsurv$trans)$transname))
table(msfB_relsurv$Haz$trans,msfB_relsurv$Haz$trans_name)

ggplot(msfB_relsurv$Haz%>%filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_wrap(~from)


### compare patients ####
ggplot(msfA_relsurv$Haz%>%filter(grepl("Death",trans_name))%>%mutate(patient="A")%>%
         rbind(msfB_relsurv$Haz%>%filter(grepl("Death",trans_name))%>%mutate(patient="B"))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=patient,lty=type))+
  geom_line()+
  facet_wrap(~from)

ggplot(msfA_relsurv$Haz%>%mutate(patient="A")%>%
         rbind(msfB_relsurv$Haz%>%mutate(patient="B"))%>%
         filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=ifelse(grepl("\\.p",trans_name),"population hazard", "excess hazard")),
       aes(x=time,y=Haz,col=patient,lty=type))+
  geom_line()+
  facet_grid(from~patient)

ggplot(msfA_relsurv$Haz%>%rbind(plot(msfA,use.ggplot=T)$data%>%select(time:trans_name))%>%
         mutate(patient="A")%>%
         rbind(msfB_relsurv$Haz%>%rbind(plot(msfB,use.ggplot=T)$data%>%select(time:trans_name))%>%
                 mutate(patient="B"))%>%
         filter(grepl("Death",trans_name))%>%
         mutate(from=case_when(grepl("Tx ",trans_name)~"Tx",
                               grepl("Rec\\+AE ",trans_name)~"Rec+AE",
                               grepl("Rec ",trans_name)~"Rec",
                               grepl("AE ",trans_name)~"AE"),
                type=case_when(grepl("\\.p",trans_name)~"population hazard",
                               grepl("\\.e",trans_name)~"excess hazard",
                               T~"total hazard")),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_grid(from~patient)+
  guides(color=guide_legend(rev=T))+ggtitle("hazards for death")

ggplot(msfA_relsurv$Haz%>%mutate(df="relsurv")%>%
         rbind(plot(msfA,use.ggplot=T)$data%>%select(time:trans_name)%>%mutate(df="normal"))%>%
         mutate(patient="A")%>%
         rbind(msfB_relsurv$Haz%>%mutate(df="relsurv")%>%
                 rbind(plot(msfB,use.ggplot=T)$data%>%select(time:trans_name)%>%mutate(df="normal"))%>%
                 mutate(patient="B"))%>%
         mutate(type=case_when(grepl("\\.p",trans_name)~"population hazard",
                               grepl("\\.e",trans_name)~"excess hazard",
                               df=="normal"~"total hazard (normal)",
                               T~"total hazard (relsurv)"),
                trans_name=gsub("\\.p","",trans_name),trans_name=gsub("\\.e","",trans_name)),
       aes(x=time,y=Haz,col=type))+
  geom_line()+
  facet_grid(patient~trans_name)+
  guides(color=guide_legend(rev=T))
# problem with Rec -> Rel; because patients with relapse cannot die? Why only for this transition??
events(msebmt2)$Frequencies

# probabilities
ptB_relsurv<-probtrans(msfB_relsurv, predt=0, variance=F)
plot(ptB_relsurv, use.ggplot=T, type="separate")+
  scale_color_manual(values=rep("black",10))+theme(legend.position="none")
plot(ptB_relsurv, use.ggplot=T)

ggplot(plot(ptA,use.ggplot=T,type="separate")$data%>%select(time,prob,state)%>%mutate(type="normal")%>%
         rbind(plot(ptA_relsurv,use.ggplot=T,type="separate")$data%>%select(time,prob,state)%>%
                 mutate(type="relsurv"))%>%
         mutate(state2=ifelse(grepl("Death",state),"Death",as.character(state)),
                death=case_when(grepl("\\.p",state)~"population mortality",
                                grepl("\\.e",state)~"excess mortality",
                                T~"total")),
       aes(x=time,y=prob,col=type,lty=death))+
  geom_step(alpha=0.75)+
  facet_wrap(~state2)+
  guides(color=guide_legend(rev=T))+ggtitle("patient A")

ggplot(plot(ptB,use.ggplot=T,type="separate")$data%>%select(time,prob,state)%>%mutate(type="normal")%>%
         rbind(plot(ptB_relsurv,use.ggplot=T,type="separate")$data%>%select(time,prob,state)%>%
                 mutate(type="relsurv"))%>%
         mutate(state2=ifelse(grepl("Death",state),"Death",as.character(state)),
                death=case_when(grepl("\\.p",state)~"population mortality",
                                grepl("\\.e",state)~"excess mortality",
                                T~"total")),
       aes(x=time,y=prob,col=type,lty=death))+
  geom_step(alpha=0.75)+
  facet_wrap(~state2)+
  guides(color=guide_legend(rev=T))+ggtitle("patient B")
```

## Problem1
coxph.relsurv(Surv(Tstart, Tstop, status)~match.1 + match.2 + match.3 + match.4 + match.5 + 
                          match.6 + match.7 + match.8 + match.9 + match.10 + match.11 + match.12 + proph.1 +
                          proph.2 + proph.3 + proph.4 + proph.5 + proph.6 + proph.7 + proph.8 + proph.9 +
                          proph.10 + proph.11 + proph.12 + year1.1 + year1.2 + year1.3 + year1.4 + year1.5 +
                          year1.6 + year1.7 + year1.8 + year1.9 + year1.10 + year1.11 + year1.12 + year2.1 +
                          year2.2 + year2.3 + year2.4 + year2.5 + year2.6 + year2.7 + year2.8 + year2.9 +
                          year2.10 + year2.11 + year2.12 + agecl1.1 + agecl1.2 + agecl1.3 + agecl1.4 +
                          agecl1.5 + agecl1.6 + agecl1.7 + agecl1.8 + agecl1.9 + agecl1.10 + agecl1.11 + 
                          agecl1.12 + agecl2.1 + agecl2.2 + agecl2.3 + agecl2.4 + agecl2.5 + agecl2.6 +
                          agecl2.7 + agecl2.8 + agecl2.9 + agecl2.10 + agecl2.11 + agecl2.12+strata(trans), 
             data=msebmt2, method="breslow",
             split.transitions = c(4,7,10,12), ratetable = joinpoptab, time.format = 'days',
             rmap=list(age=age2*365.241, year=year2))
Error in coxph.relsurv(Surv(Tstart, Tstop, status) ~ match.1 + match.2 +  : 
In split.transitions you have supplied a transition for which there are no covariates in the formula.

problem is here (line 63 in coxph.relsurv):
kovs <- strsplit(av, "\\+")[[1]]
this gives:
[1] "Surv(Tstart, Tstop, status) ~ match.1 " " match.2 "                             
[3] " match.3 "                              " match.4 "                             
[5] " "
because deparse() gives:
[1] "Surv(Tstart, Tstop, status) ~ match.1 + match.2 + match.3 + match.4 + " 
 [2] "    match.5 + match.6 + match.7 + match.8 + match.9 + match.10 + "      
 [3] "    match.11 + match.12 + proph.1 + proph.2 + proph.3 + proph.4 + "     
 [4] "    proph.5 + proph.6 + proph.7 + proph.8 + proph.9 + proph.10 + "      
 [5] "    proph.11 + proph.12 + year1.1 + year1.2 + year1.3 + year1.4 + "     
 [6] "    year1.5 + year1.6 + year1.7 + year1.8 + year1.9 + year1.10 + "      
 [7] "    year1.11 + year1.12 + year2.1 + year2.2 + year2.3 + year2.4 + "     
 [8] "    year2.5 + year2.6 + year2.7 + year2.8 + year2.9 + year2.10 + "      
 [9] "    year2.11 + year2.12 + agecl1.1 + agecl1.2 + agecl1.3 + agecl1.4 + " 
[10] "    agecl1.5 + agecl1.6 + agecl1.7 + agecl1.8 + agecl1.9 + agecl1.10 + "
[11] "    agecl1.11 + agecl1.12 + agecl2.1 + agecl2.2 + agecl2.3 + "          
[12] "    agecl2.4 + agecl2.5 + agecl2.6 + agecl2.7 + agecl2.8 + agecl2.9 + " 
[13] "    agecl2.10 + agecl2.11 + agecl2.12 + strata(trans)"
and the formula stops after the first line 

solution: use deparse1():
"Surv(Tstart, Tstop, status) ~ match.1 + match.2 + match.3 + match.4 + match.5 + match.6 + match.7 + match.8 + match.9 + match.10 + match.11 + match.12 + proph.1 + proph.2 + proph.3 + proph.4 + proph.5 + proph.6 + proph.7 + proph.8 + proph.9 + proph.10 + proph.11 + proph.12 + year1.1 + year1.2 + year1.3 + year1.4 + year1.5 + year1.6 + year1.7 + year1.8 + year1.9 + year1.10 + year1.11 + year1.12 + year2.1 + year2.2 + year2.3 + year2.4 + year2.5 + year2.6 + year2.7 + year2.8 + year2.9 + year2.10 + year2.11 +      year2.12 + agecl1.1 + agecl1.2 + agecl1.3 + agecl1.4 + agecl1.5 + agecl1.6 + agecl1.7 + agecl1.8 + agecl1.9 + agecl1.10 + agecl1.11 + agecl1.12 + agecl2.1 + agecl2.2 + agecl2.3 + agecl2.4 + agecl2.5 + agecl2.6 + agecl2.7 + agecl2.8 + agecl2.9 + agecl2.10 + agecl2.11 + agecl2.12 + strata(trans)"

## problem 2
function find_trans doesn't work if >9 transitions:
kovs[grep(paste0(".", st), sapply(kovs, 
      find_trans), fixed = TRUE)]
gives character(0) for st == 10

solution: find_trans <- function(x) {
    x <- gsub(" ", "", x, fixed = TRUE)
    substr(x, sapply(gregexpr("\\.", kovs), '[', 1)+1, nchar(x)) 
  }
-> 
kovs[grep(paste0(".", st), sapply(kovs, 
      find_trans), fixed = TRUE)]
returns now "agecl1.10" "agecl2.10"

## problem 3 
probtrans semi-parametric with relsurv gives fatal error if variance=F is not specified (default = T) -> add in multi-state code that if varHaz is missing the syntax stops?

## to do
plot.msfit for relsurv

```{r eval=F, include=F}
# reinstall published versions of mstate and relsurv

```